Dynamics of innovation in stigmergic collectives

### 2023-06-15 LaTeX/draft.tex
CodeUpdates

We know that stigmergy is one of the fundamental mechanisms by which the collective coordinates
and recruits new workers. With that being the case, the agent can follow the trail $\x^*(s)$ by making
measurements along its motion trajectory. Since we are interested in understanding how and when an
agent can innovate from its innate behavior of following pheromone trails to explore the environment and find novel solutions
to the navigation problem, we assume that the agent has already traversed the trajectory $\x^*(s)$ before.
This intrinsic behavior is denoted by the policy $\pi_{\rm int}(\r)$ which provides the
appropriate angle $\theta^t$ required to follow the trail when the agent is at location
$\r^t$ after the measurment that the agent makes. We consider the trail to be a simple semi-circle
with functional form $\xst(s) = \{ a \cos(s/a), a \sin(s/a) \}$ (see Fig.~\ref{fig:schm1}$(b)$). For this particular case,
the intrinsic trail following policy can be explicitly written as
\begin{align}
    \text{Measurement: } \phi^{t} = & \ \tan^{-1} \bigg( \frac{r_y^t}{r_x^t} \bigg) + \zeta^t, \\
    \text{Action: }\theta^{t} =& \ \frac{\pi}{2} + \phi^{t}, \\
    \text{State update: } \r^{t+1} =& \ \r^t + l \ph^{t}.
\end{align}
where $\zeta^t$ is the noise/error in the agent's measurement, $l$ is the step-length
of the agent. In the limit when the step-length is small, we can write the continuous version
of the motion as $\dot{\r} = v_o \ph$. Without loss of generality, we can write this as
$\d\r/\d s = \ph$ where $s = v_o t$. Moreover when the orientation of the agent
along its trajectory is distributed as though it were a worm-like chain polymer,
the functional form of orientation distribution is
\[
    \P[\theta(s)] = \frac{1}{\mathcal{Z}}\exp{(-\beta E_b)},
\]
where we define a bending energy for the trajectory captured by the agent's orientation as $E_b[\theta(s)]~=~B/2 \int_0^{s_c} (\theta'(s)-\kappa_o)^2~\d s$, the partition
function $\mathcal{Z}~=~\int \mathcal{D}[\theta(s)] \exp\{ {-\beta E_b[\theta(s)]} \}$ and the bending
stiffness $B$ captures the constraint for the agent to move forward towards the target A (ref Fig.~\ref{fig:schm1}$(b)$), $\beta$ is the effective
temperature capturing the fluctuations in the trajectory of agent. We also have
boundary conditions $\theta(0) = \pi/2$ (due to initial orientation of the agent) and $\theta'(0) = \kappa_o = a^{-1}$ (due to curvature of the trail).
Here $s_c$ is the coordinate along the semi-circle at which the agent bifurcates towards the target location A using a new trajectory.
Our motivation is to identify this location using the Bellman equation which will be discussed later.
Now in order to evaluate the quality of the intrinsic policy we introduce the notion of value function
of this policy $\pi_{\rm int}(\r)$, denoted by $v_{\rm int}(\r)$. The value function provides the cummulative
future reward that the agent will acquire from location $\r$ at time $t$
when following the policy $\pi_{\rm int}(\r)$. This can be written as
\[
    v_{\rm int}(\r) = \E_\pi \bigg[ \sum_{k=0}^\infty \gamma^k R_{t+k+1} \bigg| \r_t = \r \bigg].
\]
We can write this in path-integral form as (with $\gamma = 1$)
\begin{align}
    v_{\rm int}(\r) =& \ \sum_{k=t}^T \int \D[\theta(s)] \P[\theta(s)] \gamma^{k-t} R_{k+1}, \\
    \approx& \ \int_{s}^{s_c} \int \D[\theta(s)] \P[\theta(s)] R(s) \ \d s
\end{align}
The reward itself is a function of the agent's azimuthal direction, $\phi(s) = \tan^{-1}(y(s)/x(s))$.
Assuming that the reward has the function form $R(\phi) = \exp(-\phi/\phi^*)$, the leading order form of 
the value function for this policy is
\[
    v_{\rm int}(\r) = \frac{\phi^*}{\kappa_o} [ \e^{-\kappa_o s/\phi^*} - \e^{-\kappa_o s_c/\phi^*}].
\]

@@

As already mentioned, the agent reaching a distance $\sigma$ (appropriately non-dimensionalized with $l_\nu$) along $\ebh_\perp$ from the end-point A
can reach the target, then the probability that an agent will reach the target is given by $\P(||r_\perp^{s^*} - r^*|| \leq \sigma)
= \sigma/\psi(s^*)$. We can now calculate the value function, $v_{\rm OU}(\r, \theta)$ as

@@

The reward function for the OU process takes a simple form $R(s) = \mu[1 - \Theta(||r_\perp(s^*) - r^*|| - \sigma)]$,

@@

>>>>

We know that stigmergy is one of the fundamental mechanisms by which the collective coordinates
and recruits new workers. With that being the case, the agent can follow the trail $\x^*(s)$ by making
measurements along its motion trajectory. Since we are interested in understanding how and when an
agent can innovate from its innate behavior of following pheromone trails to explore the environment and find novel solutions
to the navigation problem, we assume that the agent has already traversed the trajectory $\x^*(s)$ before.
This intrinsic behavior is denoted by the policy $\pi_{\rm int}(\r)$ which provides the
appropriate angle $\theta^t$ required to follow the trail when the agent is at location
$\r^t$ after the measurment that the agent makes. We consider the trail to be a simple semi-circle
with functional form $\xst(s) = \{ a \cos(s/a), a \sin(s/a) \}$ (see Fig.~\ref{fig:schm1}$(b)$). For this particular case,
the intrinsic trail following policy can be explicitly written as
\begin{align}
    \text{Measurement: } \phi^{t} = & \ \tan^{-1} \bigg( \frac{r_y^t}{r_x^t} \bigg) + \zeta^t, \\
    \text{Action: }\theta^{t} =& \ \frac{\pi}{2} + \phi^{t}, \\
    \text{State update: } \r^{t+1} =& \ \r^t + l \ph^{t}.
\end{align}
where $\zeta^t$ is the noise/error in the agent's measurement, $l$ is the step-length
of the agent. In the limit when the step-length is small, we can write the continuous version
of the motion as $\dot{\r} = v_o \ph$. Without loss of generality, we can write this as
$\d\r/\d s = \ph$ where $s = v_o t$. Moreover when the orientation of the agent
along its trajectory is distributed as though it were a worm-like chain polymer,
the functional form of orientation distribution is
\[
    \P[\theta(s)] = \frac{1}{\mathcal{Z}}\exp{(-\beta E_b)},
\]
where we define a bending energy for the trajectory captured by the agent's orientation as $E_b[\theta(s)]~=~B/2 \int_0^{s_c} (\theta'(s)-\kappa_o)^2~\d s$, the partition
function $\mathcal{Z}~=~\int \mathcal{D}[\theta(s)] \exp\{ {-\beta E_b[\theta(s)]} \}$ and the bending
stiffness $B$ captures the constraint for the agent to move forward towards the target A (ref Fig.~\ref{fig:schm1}$(b)$), $\beta$ is the effective
temperature capturing the fluctuations in the trajectory of agent. We also have
boundary conditions $\theta(0) = \pi/2$ (due to initial orientation of the agent) and $\theta'(0) = \kappa_o = a^{-1}$ (due to curvature of the trail).
Here $s_c$ is the coordinate along the semi-circle at which the agent bifurcates towards the target location A using a new trajectory.
Our motivation is to identify this location using the Bellman equation which will be discussed later.
Now in order to evaluate the quality of the intrinsic policy we introduce the notion of value function
of this policy $\pi_{\rm int}(\r)$, denoted by $v_{\rm int}(\r)$. The value function provides the cummulative
future reward that the agent will acquire from location $\r$ at time $t$
when following the policy $\pi_{\rm int}(\r)$. This can be written as
\[
    v_{\rm int}(\r) = \E_\pi \bigg[ \sum_{k=0}^\infty \gamma^k R_{t+k+1} \bigg| \r_t = \r \bigg].
\]
We can write this in path-integral form as (with $\gamma \approx 1$)
\begin{align}
    v_{\rm int}(\r) =& \ \sum_{k=t}^T \int \D[\theta(s)] \P[\theta(s)] \gamma^{k-t} R_{k+1}, \\
    \approx& \ \int_{s}^{s_c} \int \D[\theta(s)] \P[\theta(s)] R(s) \ \d s
\end{align}
The reward itself is a function of the agent's azimuthal direction, $\phi(s) = \tan^{-1}(y(s)/x(s))$.
Assuming that the reward has the function form $R(\phi) = \exp(-\phi/\phi^*)$, the leading order form of 
the value function for this policy is
\[
    v_{\rm int}(\r) = \frac{\phi^*}{\kappa_o} [ \e^{-\kappa_o s/\phi^*} - \e^{-\kappa_o s_c/\phi^*}].
\]

@@

As already mentioned, the agent reaching a distance $\sigma$ (appropriately non-dimensionalized with $l_\nu$) from the end-point A
can reach the target, then the probability that an agent will reach the target is given by $\P(||\r(s) - \r^*|| \leq \sigma)
= 2\int_0^\sigma \mu \sqrt{\sigma^2-\zeta^2} \ \d \zeta/\psi(s^*-\zeta)$. We can now calculate the value function, $v_{\rm OU}(\r, \theta)$ as

@@

% The reward function for the OU process takes a simple form $R(s) = \mu[1 - \Theta(||r_\perp(s^*) - r^*|| - \sigma)]$,
The reward function for the OU process takes a simple form $R(s) = \mu[1 - \Theta(||\r(s^*) - \r^*||^2 - \sigma^2)]$,

@@

Mention the limits of $\sigma$ set by the criteria that $\psi(s) > 0$.
